---
title: "Cancer EDA"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Setup

First, we load the car library, which gives us a convenient scatterplotMatrix function.

```{r car}
library(car)
```

```{r loading}
# Load the data 
cancer.df=read.csv("cancer.csv")
```

## Data Transformation

We're going to explore a set of variables that represent the levels of health insurance coverage for individual counties.There are three variables in the original dataset that are related to insurance:

Variable Name         Description    
-------------------   -----------------------------------------------------------------------  
**PctPrivateCoverage**    Percentage of the population with private insurance coverage
**PctPublicCoverage**     Percentage of the population with public insurance coverage
**PctEmpPrivCoverage**    Percentage of the population with employer-sponsored private insurance coverage
-------------------   -----------------------------------------------------------------------

For the purposes of our explanatory analysis, we would like to conduct a more comprehensive research on various types and levels of insurance coverage and their effects on the mortality rates, so it makes sense to define a few more variables that can be derived from the original dataset.
For example, we would like to include data about the populations with no insurance coverage, as well as the observations where individuals have both private and public insurance. It can also be more revealing to treat the employer-sponsored coverage as a relative proportion of the private coverage rather than an absolute value.

Hence, let's introduce three new variables as follows:

Variable Name          Description    
-------------------    -----------------------------------------------------------------------  
**PctPNoCoverage**     Percentage of the population with no insurance coverage
**PctDoubleCoverage**  Percentage of the population with both private and public insurance coverage
**EmpSponsoredPct**    Percentage of the private insurance sponsored by employers
-------------------    -----------------------------------------------------------------------

We will now add these new variables to our original dataset:

```{r transformation}
nrow(cancer.df[(cancer.df$PctPublicCoverage + cancer.df$PctPrivateCoverage)>100,])
nrow(cancer.df[(cancer.df$PctPublicCoverage + cancer.df$PctPrivateCoverage)<100,])

cancer.df$PctDoubleCoverage=cancer.df$PctPublicCoverage + cancer.df$PctPrivateCoverage - 100
cancer.df$PctDoubleCoverage[cancer.df$PctDoubleCoverage < 0] = 0
summary(cancer.df$PctDoubleCoverage)

cancer.df$PctNoCoverage = 100 - cancer.df$PctPublicCoverage - cancer.df$PctPrivateCoverage
cancer.df$PctNoCoverage[cancer.df$PctNoCoverage < 0] = 0
summary(cancer.df$PctNoCoverage)

cancer.df$EmpSponsoredPct = cancer.df$PctEmpPrivCoverage / cancer.df$PctPrivateCoverage * 100
summary(cancer.df$EmpSponsoredPct)
```

## Univariate Analysis of Key Variables

Our key variables in this investigation will be deathRate (target variable) and several indpendent variables representing insurance coverage for counties' populations.

### Cancer Mortality Rate (**deathRate** variable)
Let's start with the target variable and summarize it:


```{r deathRate}
summary(cancer.df$deathRate)
```
We see that this is a metric variable with its mean and median values very close to each other. There are no missing values and no obviously wrong or suspicious outliers.

To better visualize the variable's values distribution, we plot the histogram.

```{r}
with(cancer.df, hist(deathRate,  col = "gray", 
                     main="Histogram of Cancer Death Rates",
                     xlab="death rate"))
box()
```
As we can see from the output, the default method for selecting the number of bins produced too few bins, which might obscure some interesting features in the data. A better result is achieved by setting the binning rule to the one proposed by Freedman and Diaconis. Fortunately, hist() function has a built-in option for this:
```{r}
with(cancer.df, hist(deathRate,  breaks='FD', col = "gray", 
                     main="Histogram of Cancer Death Rates",
                     xlab="death rate"))
box()
```
Now we have a much higher level of detail and can easily infer that deathRate variable distribution is very close to the normal one, with a notable outliers on the far right of the histogram.

Let's explore the extreme outliers with deathRate over 300 and see if we can find anything unusual in these observations.
To find out how many outliers are there, we'll use the nrow() function:
```{r}
nrow(cancer.df[cancer.df$deathRate > 300,])
```
Turns out there's only one observation with this property, so let's examine it a bit closer.


```{r}
str(cancer.df[cancer.df$deathRate > 300,])
```

At first sight, nothing in the rest of the data stands out to provide a possible explanation for the high mortality rate (363). We might want to revisit this observation once we completed the rest of the analysis.


### Private Insurance Coverage (**PctPrivateCoverage** variable)

Similar to our target variable, we summarize PctPrivateCoverage and generate its histogram: 

```{r PctPrivateCoverage}
summary(cancer.df$PctPrivateCoverage)
with(cancer.df, hist(PctPrivateCoverage, breaks="FD", col = "gray", 
                     main="Histogram of Private Insurance Coverage",
                     xlab="percent covered"))
box()
```
We notice that the frequency distribution has some negative skew, with the majority of values falling between 55% and 75%. The data looks reasonable, with no obvious errors and missing values.

### Public Insurance Coverage (**PctPublicCoverage** variable)

We repeat the steps executed above for the public insurance coverage:

```{r PctPublicCoverage}
summary(cancer.df$PctPublicCoverage)
with(cancer.df, hist(PctPublicCoverage, breaks="FD", col = "gray", 
                     main="Histogram of Public Insurance Coverage",
                     xlab = "percent covered"))
box()
```

Compared to the private insurance coverage, the data is more evenly distributed and is much closer to the normal curve. The mean and median values are almost half of the ones for the private insurance coverage. From that we can infer that the private insurance is much more prevalent than the one sponsored by the state.
Similar to PctPrivateCoverage, the public coverage variables doesn't show any obvious errors and there are no missing values.

### Employer-sponsored portion of the private coverage (**EmpSponsoredPct** variable)

After exploring the general category of the private coverage, we would like to examine what portion of the insurance are provided by employers:

```{r EmpSponsoredPct}
summary(cancer.df$EmpSponsoredPct)
with(cancer.df, hist(EmpSponsoredPct, breaks="FD", col = "gray", 
                     main="Histogram of Employer Portion of Private Coverage",
                     xlab = "percent covered"))
box()
```
The histogram tells us that employment is the major source of private insurance coverage in the counties: most of the values of EmpSponsoredPct variable fall between 60% and 70%.


### No insurance coverage (**PctNoCoverage** variable)
Let's summarize our generated variable that represents percentage of the population with no insurance coverage:

```{r PctNoCoverage}
summary(cancer.df$PctNoCoverage)
with(cancer.df, hist(PctNoCoverage, breaks="FD", col = "gray", 
                     main="Histogram of No Insurance Coverage",
                     xlab = "percent not covered"))
box()
```
Unlike the distributions we've seen so far, this variable has a major peak around 0, with the rest of the values tapering off in the shape of the long-tailed distribution. 
To get a better insight into the variable, we can generate the percentile metric:

```{r}
quantile(cancer.df$PctNoCoverage, prob = seq(0, 1, length = 11), type = 5)
```
The result shows that 80% of the observations have less than 5% of the population with no health insurance. We can safely infer then that the effect of this variable on the target will be minimal.

### Coverage that includes both private and public components (**PctDoubleCoverage** variable)

We repeat the steps executed during the evaluation of PctNoCoverage variable:

```{r PctDoubleCoverage}
summary(cancer.df$PctDoubleCoverage)


with(cancer.df, hist(PctDoubleCoverage, breaks="FD", col = "gray", 
                     main="Histogram of Double Coverage",
                     xlab = "percent covered"))
box()

quantile(cancer.df$PctDoubleCoverage, prob = seq(0, 1, length = 11), type = 5)
```
The result shows that 80% of the counties have less than 7% of the population with double health insurance. Therefore, similar to the previous case, its relative effect on the target variable will be minimal.

## Analysis of Key Relationships

### Mortality rates for different levels of private insurance coverage

Our first question is whether having access to a private insurance coverage is correlated with cancer mortality rates. A reasonable hypothesis would be that a cancer patient with a private insurance would be able to afford better treatment options. As a result, she or he will have better chances of survival, so we should expect negative correlation between deathRate and PctPrivateCoverage.
Let's build a scatterplot showing the relationshoip between these two variables. 
In order to get a better insight into what linear relationship exists in the data, we add the ordinary least squares 
regression line to the plot and calculate the correlation.

```{r deathRate_PctPrivateCoverage }
plot(cancer.df$PctPrivateCoverage, cancer.df$deathRate, 
     xlab = "percent with private coverage", ylab = "death rate", 
     main = "Death rates for different levels of private insurance coverage")
abline(lm(cancer.df$deathRate ~ cancer.df$PctPrivateCoverage))

cor(cancer.df$deathRate, cancer.df$PctPrivateCoverage)
```
Both from the plot and from the correlation value (-0.39) we can see that they're in agreement with our original hypothesis that mortality rates ared lower for the populations with higher percentage of private insurance coverage. The relationship does appear to be linear from about 40% of coverage onward (this is where the majority of observations seem to fall). At the lower end of the graph, the spread of values is much higher.
Despite showing the overall trend, the scatterplot is quite noisy, so we might want to confirm our conclusion by generating boxplots for different categories of coverage. 
First, we'll split the range of PctPrivateCoverage variables into three bins and label them as "low", "medium", and "high" brackets of private insurance coverage. We then will build three separate boxplots for these categories and see how they're distributed relative to deathRate.
```{r}
levels(cut(cancer.df$PctPrivateCoverage, 3, include.lowest=TRUE))
boxplot(deathRate ~ cut(PctPrivateCoverage, 3, include.lowest=TRUE, 
        labels=c("low", "medium", "high")), 
        data = cancer.df,
        cex.axis = .7,
        main = "Death Rate for different levels of private insurance coverage",
        xlab = "private insurance coverage", ylab = "death rate")
```
The boxplot shows a clear downward trend from the "medium" to "high" category, with the majority of values clustered around the median. The "low" category boxplot, on the other hand, has a much wider spread of data points. We might conclude, therefore, that the effect of private insurance on mortality rates is only noticable for the percantage of coverage which is above certain threshold (~40%).
The "medium" category also includes the high death rate outlier we've identified earlier (>350). Therefore, the high mortality rate can't be explained by the inadequate private insurance coverage.

__Summary of observations:__

1. There's a mild negative correlation between cancer mortality rates and access to the private insurance coverage

2. The effect of negative correlation becomes noticable only after the coverage percentage reaches ~40%. Below this point, the data spread is much wider and the effect of private coverage is not obvious.


### Mortality rates for different levels of public insurance coverage

We now explore whether public insurance coverage has a similar effect on cancer mortality rates. We repeat the same steps of data analysis we've performed for the private insurance variable:
```{r deathRate_PctPublicCoverage}
plot(cancer.df$PctPublicCoverage, cancer.df$deathRate,  
     xlab = "percent with public coverage", ylab = "death rate", 
     main = "Death rates for different levels of public insurance coverage")
abline(lm(cancer.df$deathRate ~ cancer.df$PctPublicCoverage))

cor(cancer.df$deathRate, cancer.df$PctPublicCoverage)
```


```{r}
levels(cut(cancer.df$PctPublicCoverage, 3, include.lowest=TRUE))
boxplot(deathRate ~ cut(PctPublicCoverage, 3, include.lowest=TRUE, 
        labels=c("low", "medium", "high")), 
        data = cancer.df,
        cex.axis = .7,
        main = "Death Rate for different levels of public insurance coverage",
        xlab = "public insurance coverage", ylab = "death rate")
```
Contrary to our expectations, we see the directly opposite relationship between public insurance coverage and cancer mortality rates. The values are positively correlated and the correlation's absolute value is even higher than the one we calculated for private insurance coverage.
There's also no salient threshold effect we observed earlier: the relationship appears to be linear throughout the entire range of coverage percentage.


__Summary of observations:__

1. There's a noticeble positive correlation between cancer mortality rates and availability of public insurance coverage

2. The relationship is very close to the linear one throughout the entire range of coverage's percentages


### Relationship between private and public insurance coverage

We will now explore if there is any meaningful relationship between private and public insurance coverage.
As in the earlier steps of our investigation, we generate a scatterplot and box plots for these variables, and compute the correlation value:

```{r PctPrivateCoverage_PctPublicCoverage}

plot(cancer.df$PctPublicCoverage, cancer.df$PctPrivateCoverage,  
     xlab = "percent with public coverage", ylab = "percent with private coverage", 
     main = "Private coverage for different levels of public insurance coverage")
abline(lm(cancer.df$PctPrivateCoverage ~ cancer.df$PctPublicCoverage))

cor(cancer.df$PctPrivateCoverage, cancer.df$PctPublicCoverage)
```
```{r}
levels(cut(cancer.df$PctPublicCoverage, 3, include.lowest=TRUE))
boxplot(PctPrivateCoverage ~ cut(PctPublicCoverage, 3, include.lowest=TRUE, 
        labels=c("low", "medium", "high")), 
        data = cancer.df,
        cex.axis = .7,
        main = "Private coverage for different levels of public insurance coverage",
        xlab = "public insurance coverage", ylab = "percent with private coverage")


```

__Summary of observations:__

1. There's a strong negative correlation between private and public insurance coverage

2. The majority of observations cluster around ordinary least squares regression line, emphasizing linear relationship between the two variables



### Mortality rates for different levels of employer-sponsored private coverage

Finally, let's see if the relative portion of employer-sponsored private insurance coverage has any relationship 
with cancer mortality rates.

```{r deathRate_EmpSponsoredPct}
plot(cancer.df$EmpSponsoredPct, cancer.df$deathRate,  
     xlab = "percentage of private coverage provided by employers", 
     ylab = "death rate", 
     main = "Death rates for different levels of employer coverage")
abline(lm(cancer.df$deathRate ~ cancer.df$EmpSponsoredPct))

cor(cancer.df$deathRate, cancer.df$EmpSponsoredPct)
```



__Summary of observations:__

1. From the data analysis above, we don't detect any noticable relationships between the cancer mortality rates and the composition of the private insurance coverage.



### Conclusion: insurance coverage per ce doesn't improve cancer mortality rates

1. The data analysis we performed has refuted our hypothesis that cancer patients who have access to health insurance have better chances of survival. 

2. We also saw that private and public insurance demonstrate opposite relationships with cancer mortality rates.

3. One of the important findings is that higher levels of public insurance coverage are strongly correlated with lower percentage of private insurance coverage. 

4. The relative amount of private coverage sponsored by employers have no detectable relationships with cancer mortality rates.

5. In order to explain these counter-intuitive results, as well as the 'threshold effect' of private coverage, we need to explore other variables that might directly influence these relationships.

## Analysis of Secondary Effects

Since we have seen that private and public insurance have opposite relationships with cancer mortality rates and highly negatively correlated between each other, we now must explore other variables that could be influencing these relationships.

We will start with a univariate analysis of other selected variables

_avgAnnCount_
```{r avgAnnCount}
hist(cancer.df$avgAnnCount, breaks="fd", xlab = "Mean Number of Incidences per County (2009-2013)", ylab="Frequency", main = "Histogram of Mean Cancer Incidences", col= "gray")
```
That second peak seems a little odd. We should take a look at what is going on there, where the avgAnnCount is between 1500 and 2500

```{r outliers}
hist(cancer.df$avgAnnCount[cancer.df$avgAnnCount>1500 & cancer.df$avgAnnCount<2500],xlab="avgAnnCount between 1500 and 2500", main = "Histogram of avgAnnCount outliers",col="gray")
```

We can now see that the peaks are between avgAnnCount of 1900 and 2000, a range easier to investigate:

```{r outliers investigation}
cancer.df[cancer.df$avgAnnCount>1900 & cancer.df$avgAnnCount<2000,]
```

We notice that there are a lot of observations that have the exact same value of avgAnnCount - 1962.667684 - and this is the only not integer value in the whole set of avgAnnCount values. For that matter, before any further analysis, we will transform all of these values in NAs.

```{r correcting avgAnnCount}
cancer.df$avgAnnCount[cancer.df$avgAnnCount==1962.667684] = NA
```

Now we can go back to analyzing the avgAnnCount:

```{r avgAnnCount}
hist(cancer.df$avgAnnCount, breaks="fd", xlab = "Mean Number of Incidences per County (2009-2013)", ylab="Frequency", main = "Corrected Histogram of Mean Cancer Incidences", col= "gray")
```
That extremely right-skewed distribution is an indicative that we could use a log() transformation in this variable. Let's see how the variable is before the transformation.

```{r Further analysis}
summary(cancer.df$avgAnnCount)
```
Now to the analysis of the log(avgAnnCount)

```{r log(avgAnnCount)}
summary(log(cancer.df$avgAnnCount))
```

```{r log(avgAnnCount)}
hist(log(cancer.df$avgAnnCount), breaks="fd", xlab = "log(avgAnnCount)", ylab="Frequency", main = "Histogram of log(avgAnnCount)", col= "gray")
```

The distribution of log(avgAnnCount) is fairly close to normal.

_medIncome_
```{r medIncome}
hist(cancer.df$medIncome, breaks="fd", xlab = "Median Income", ylab="Frequency", main = "Histogram of Median Income", col= "gray")
```

The distribution of Median Income is fairly normal with a slight right skew.

_povertyPercent_
```{r povertyPercent}
hist(cancer.df$povertyPercent, breaks="fd", xlab = "Percentage of population below poverty line", ylab="Frequency", main = "Histogram of Poverty Percentage", col= "gray")
```

The distribution of Poverty Percent is fairly normal with a slight right skew.

# Bivariate Analysis of Secondary Variables

After taking a first look into other variables, and performing necessary corrections, we must now understand how they relate to the primary key variables, in order to comprehend what else might be driving the relationship previously found between different types of health insurance coverage and cancer death rates.

For each secondary variable introduced, we will analyze their relationship with the primary variables and with the output variable itself.

_log(avgAnnCount)_

Private Insurance Coverage

```{r log(avgAnnCount) vs Private Insurance Coverage}
plot(log(cancer.df$avgAnnCount),cancer.df$PctPrivateCoverage, ylab = "% population with private health insurance", xlab="log(avgAnnCount)", main= "Private Coverage vs log(avgAnnCount)")
abline(lm(cancer.df$PctPrivateCoverage[!is.na(cancer.df$avgAnnCount)] ~ log(cancer.df$avgAnnCount[!is.na(cancer.df$avgAnnCount)])))

cor(log(cancer.df$avgAnnCount),cancer.df$PctPrivateCoverage, use = "complete.obs")
```
There is a small positive correlation between the log(avgAnnCount) and the percentage of population with private health insurance coverage.

Public Insurance Coverage

```{r log(avgAnnCount) vs Public Insurance Coverage}
plot(log(cancer.df$avgAnnCount),cancer.df$PctPublicCoverage, ylab = "% population with public health insurance", xlab="log(avgAnnCount)", main= "Public Coverage vs log(avgAnnCount)")
abline(lm(cancer.df$PctPublicCoverage[!is.na(cancer.df$avgAnnCount)] ~ log(cancer.df$avgAnnCount[!is.na(cancer.df$avgAnnCount)])))

cor(log(cancer.df$avgAnnCount),cancer.df$PctPublicCoverage, use = "complete.obs")
```

There is a small negative correlation between the log(avgAnnCount) and the percentage of population with public health insurance coverage, in a very close magnitude to the positive correlation encountrerd with the private health insurance coverage.We have seen before that, in terms of death rate, these two types of health insurance coverage have opposite behaviors. That being said, even if it is a small correlation, the fact that it presents itself in opposite ways and in similar magnitude to public and private health insurance coverage, just like death rates, indicates that we should dig deeper to check if the log(avgAnnCount) has a stronger positive correlation with death rate.

Death Rate
```{r log(avgAnnCount) vs Death Rate}
plot(log(cancer.df$avgAnnCount),cancer.df$deathRate, ylab = "Cancer Death Rate", xlab="log(avgAnnCount)", main= "Death Rate vs log(avgAnnCount)")
abline(lm(cancer.df$deathRate[!is.na(cancer.df$avgAnnCount)] ~ log(cancer.df$avgAnnCount[!is.na(cancer.df$avgAnnCount)])))

cor(log(cancer.df$avgAnnCount),cancer.df$deathRate, use = "complete.obs")
```

At first sight, by just analyzing these charts and the correlation, it seems as we don't have much of a relation between these two variables. However, since they presented the same opposite behavior with public and private health insurance coverage, we want to take a deeper look into what might be going on.

```{r boxplot}

levels(cut(log(cancer.df$avgAnnCount), 7, include.lowest=TRUE))
boxplot(deathRate ~ cut(log(avgAnnCount), 7, include.lowest=TRUE), 
        data = cancer.df,
        cex.axis = .7,
        main = "Death Rate for different levels of incidence rate",
        xlab = "log(avgAnnCount)", ylab = "Death Rate")
```

With this visualization it seems as the incidence has a positive correlation with the death rate up to a certain point. But passed that threshold (4.29, 5.54], the correlation becomes negative. One possible interpretation we had is that after a certain number of cancer reported cases, there is a more pressing need to invest in that disease, increasing the survival chances for those with it and, consequently, decreasing the death rates.

_medIncome_

Private Insurance Coverage

```{r log(avgAnnCount) vs Private Insurance Coverage}
plot(cancer.df$medIncome,cancer.df$PctPrivateCoverage, ylab = "% population with private health insurance", xlab="Median Income", main= "Private Coverage vs Median Income")
abline(lm(cancer.df$PctPrivateCoverage ~ cancer.df$medIncome))

cor(cancer.df$medIncome,cancer.df$PctPrivateCoverage)
```
By the chart presented and the strong positive correlation we atest something probably intuitevely known: populations with higher income tend to have more private health insurance coverage. We can take a deeper look by analyzing boxplots for different levels of median income:


```{r boxplot}

boxplot(cancer.df$PctPrivateCoverage ~ cut(cancer.df$medIncome, right=FALSE,seq(0,160000,20000),labels = c("0 - 20k","20k - 40k","40k - 60k","60k - 80k", "80k - 100k", "100k - 120k", "120k - 140k","140k +")),
        main = "Private Coverage for different levels of income",
        xlab = "Median Income", ylab = "Private Coverage")
```

It confirms what we previously stated. There is a clear correlation between wealth and private health insurance coverage