---
title: "Exploratory Data Analysis - Incidents of Cancer"
author: "Jonathan D'Souza"
date: "9/15/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
```

## Introduction
Given a Data Set for cancer incidences for a select group of counties.... this study attempts to explore the relationships between the outcome variable : Annual Incident Count and other key independent variables.

------


```{r LoadData}
cancer<-read.csv("cancer.csv")  #Assumes file in current working directory
names(cancer)
nrow(cancer)
```
Annual Incident Count is  better expressed as percent of population

```{r CleanData}
summary(cancer$avgAnnCount)
cancer$AnnCountPercent<-with(cancer,100*avgAnnCount/popEst2015)

```





## Univariate Analysis of Key Variables
It is clear that the annual count percent has some outliers given that max % >100 (can't be more incidents than the population)
Plotting the Avg annual count shows a big spike in values
```{r univariate}
summary (cancer$AnnCountPercent)

hist(cancer$avgAnnCount,100000)

#Try with smaller range
hist(cancer$avgAnnCount,100000,xlim=c(1900,2010))
#Get these outlier values

cleandata<-subset(cancer,avgAnnCount>1970 & avgAnnCount>1960)
hist(cleandata$avgAnnCount,100000)

```


## Analysis of Key Relationships
Explore how your outcome variable is related to the other variables in your dataset. Make sure to use visualizations to understand the nature of each bivariate relationship.
What transformations can you apply to clarify the relationships you see in the data? Be sure to justify each transformation you use.

```{r relationships, echo=FALSE}
# Seems to be a correlation between death rate and incidents and a negative correlation between birth rate and incidents
scatterplotMatrix(~AnnCountPercent+popEst2015+BirthRate+deathRate,data=cleandata)

#Seems to be positive correlation between percentage of private coverage and incidents AS well as % of public coverage and incidents...? 
scatterplotMatrix(~AnnCountPercent+PctPrivateCoverage+PctEmpPrivCoverage+PctPublicCoverage,data=cleandata)

scatterplotMatrix(~AnnCountPercent+PctEmployed16_Over+PctUnemployed16_Over,data=cleandata)
```




## Analysis of Secondary Effects (10 pts)
What secondary variables might have confounding effects on the relationships you have identified? Ex- plain how these variables affect your understanding of the data.


##Conclusion (20 pts)

Summarize your exploratory analysis. What can you conclude based on your analysis? 2



