---
title: "Exploratory Data Analysis - Incidents of Cancer"
authors: "Lina Gurevich,Duda Espindola, Jonathan D'Souza "
date: "9/26/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
```

## Introduction
Given a Data Set for cancer incidences for a select group of counties.... this study attempts to explore the relationships between the outcome variable : Annual Incident Count and other key independent variables.

------


```{r LoadData}
raw_data<-read.csv("cancer.csv")  #Assumes file in current working directory
cancer<-raw_data #Keep one copy of raw data as is
str(cancer)
```
There are 31 variables across 3047 observations in this data set

```{r Univariate Analysis}
summary(cancer)
Emp.UnEmp<-cancer$PctEmployed16_Over+cancer$PctUnemployed16_Over
summary(Emp.UnEmp)
hist(Emp.UnEmp,breaks="fd",main="Distribution of Employment data per county",xlab="Percent of 16_over employed and unemployed")
```
There are 2 variables with null values: PctSomeCol18_24 and PctEmployed16_Over
The sum of the variables percentage employed and unemployed over 16 has a surprisingly broad distribution around the mean of 62.01, when one would expect it to be close to (if not) 100%. 
We will keep those aside and look at other variables

```{r AnalyzeOutcomeVariables}

#Annual Indident Rate is better expressed as a percentage of county population
cancer$AnnCountPercent<-with(cancer,100*avgAnnCount/popEst2015)
summary(cancer$AnnCountPercent)
#Look for where the outlier might be coming from

hist(cancer$avgAnnCount,breaks="fd",main="Average Annual Count Distrubution",xlab="Average Annual Count",xlim=c(0,10000))
outliers<-cancer[cancer$AnnCountPercent>50,] #Assuming anything over 50% incident rate has to be an error
summary(outliers$avgAnnCount)
#Clearly all of these have the exact same erroneous value for Average Annual Count.
error_value<-outliers[1,"avgAnnCount"]
cancer$avgAnnCount[cancer$avgAnnCount==error_value]<-NA
summary(cancer$avgAnnCount)
cancer$AnnCountPercent<-with(cancer,100*avgAnnCount/popEst2015) #Recalculate percentages
summary(cancer$AnnCountPercent)

summary(cancer$deathRate)
hist(cancer$deathRate,breaks="fd",main="Death Rate Distrubution",xlab="Death Rate")


scatterplot(AnnCountPercent~deathRate,data=cancer)
```





## Prep final data set for analysis
It is clear that the annual count percent has some outliers given that max % >100 (can't be more incidents than the population)
Plotting the Avg annual count shows a big spike in values


hist(cancer$avgAnnCount,100000)

#Try with smaller range
hist(cancer$avgAnnCount,100000,xlim=c(1900,2010))
#Get these outlier values

cleandata<-subset(cancer,avgAnnCount>1970 & avgAnnCount>1960)
hist(cleandata$avgAnnCount,100000)

```


## Analysis of Key Relationships
Explore how your outcome variable is related to the other variables in your dataset. Make sure to use visualizations to understand the nature of each bivariate relationship.
What transformations can you apply to clarify the relationships you see in the data? Be sure to justify each transformation you use.

```{r relationships, echo=FALSE}
# Seems to be a correlation between death rate and incidents and a negative correlation between birth rate and incidents
#scatterplotMatrix(~AnnCountPercent+popEst2015+BirthRate+deathRate,data=cleandata)

#Seems to be positive correlation between percentage of private coverage and incidents AS well as % of public coverage and incidents...? 
#scatterplotMatrix(~AnnCountPercent+PctPrivateCoverage+PctEmpPrivCoverage+PctPublicCoverage,data=cleandata)

#scatterplotMatrix(~AnnCountPercent+PctEmployed16_Over+PctUnemployed16_Over,data=cleandata)
```




## Analysis of Secondary Effects (10 pts)
What secondary variables might have confounding effects on the relationships you have identified? Ex- plain how these variables affect your understanding of the data.


##Conclusion (20 pts)

Summarize your exploratory analysis. What can you conclude based on your analysis? 2



